---
title: "Literate Analysis with Ocean Water Samples"
subtitle: "coreR Session 05 — Section 5.7 Practice"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This Quarto document follows the **Section 5.7 practice exercise** from coreR *Session 05 (Literate Analysis with Quarto)*: building a simple **prose + code** analysis using real seawater chemistry data (`BGchem2008data.csv`).\
Source: <https://learning.nceas.ucsb.edu/2024-10-coreR/session_05.html>

## About the data

The seawater chemistry dataset used here is `BGchem2008data.csv` (downloaded previously in the course).

-   Date downloaded: August 8, 2025
-   Download source: <https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A35ad7624-b159-4e29-a700-0c0770419941>

You will explore the nitrogen to phosphate ratio, which indicates the health of the ocean ecosystem. <https://beal-agulhas.earth.miami.edu/teaching/courses/lecture-five/index.html>

The Redfield Ratio is the ideal ratio of Nitrogen:Phosphate within an ocean's ecosystem for phytoplankton to flourish. Phytoplankon are important to maintaining the balance of a particular ocean's ecosystem. If the ratio deviates from the ideal 16:1, it indicates that there are changes in the ocean's regular functioning from upwelling, climate change/pollution, or unusual ocean currents. The data used for this project is from Arctic oceans, and can be compared to the ideal Redfield ratio. This can indicate if the Arctic ocean is experiencing changes in its ecosystem from either internal or external oceanic changes or events.

## Setup

Best practice: put all `library()` calls in one chunk near the top so collaborators can see what packages are required.

```{r}
library(here)
library(readr)
```

# Read in the data

Quarto runs R in the same way as a script: objects only exist after the chunks that create them have been run.

Using `here()` is a convenient way to build paths relative to your project root (e.g., your `.qmd` might live in `scripts/` while data lives in `data/`).

```{r}
bg_chem <- read_csv(here::here("data/raw", "BGchem2008data.csv"))
```

(Optional) Quick peek at the data:

```{r}
# Uncomment if you want more inspection output
 names(bg_chem)
# str(bg_chem)
# head(bg_chem)
# summary(bg_chem)
```

# Analysis

As the “analysis”, we’ll calculate a few simple summary statistics and examine how the nitrogen:phosphate values compare to the **Redfield ratio** concept.

## Calculate summary statistics

Compute mean nutrient values (using `na.rm = TRUE` so missing values don’t break the calculation).

```{r}
nitrate <- mean(bg_chem$NO3, na.rm = TRUE)
nitrite <- mean(bg_chem$NO2, na.rm = TRUE)
amm     <- mean(bg_chem$NH4, na.rm = TRUE)
phos    <- mean(bg_chem$P,   na.rm = TRUE)
```

## Calculate mean Redfield ratio

```{r}
ratio <- (nitrate + nitrite + amm) / phos
ratio
```

You can report results inline in Quarto like this:

The Redfield ratio for this dataset is approximately: `r round(ratio)`

## Plot Redfield ratio measurements

Finally, create a simple base R plot that shows the relationship between phosphate and the total nitrogen components.

```{r}
plot(
  bg_chem$P,
  bg_chem$NO2 + bg_chem$NO3 + bg_chem$NH4,
  xlab = "Phosphate (P)",
  ylab = "Total nitrogen (NO2 + NO3 + NH4)",
  main = "Ocean water samples: total N vs P"
)
```

This plot is a visual representation of the Redfield ratio in the Arctic Ocean. Calculated above was the Redfield ratio of 6:1 of N:P. This number is an average of all the data and should be interpreted as such. The plot tells a different story. It appears that with increasing ratio amounts of phosphate, there is also an increase in nitrogen. This follows a linear trend with the exceptions of a few outliers appearing at one on the phospate axis. The linear trend shows that when the amount of nitrogen approaches 16, the amount of phosphate is almost 2. This is significantly deviated from the ideal ration of 16:1, and poses that at this point the ratio is closer to 8:1. It can be interpreted from the plot that the Arctic ocean is lacking a source of nitrogen from some ecological, environmental, or climate impact.

# Conclusion

In this mini-analysis, we:

-   Loaded packages in a dedicated setup chunk
-   Read in `BGchem2008data.csv` using a project-root-relative path (`here()`)
-   Calculated simple nutrient means
-   Computed a mean nitrogen:phosphate ratio
-   Created a basic plot relating total nitrogen to phosphate

This is the core idea of **literate analysis**: combine narrative and code in one document so the workflow is easier to understand and reproduce.
