---
title: "a03"
author: "Sam Samler"
date: "`r Sys.Date()`"
format: html
---

# Data Analysis of Alaska Commercial Fishing from 1886-1997

## Loading Data Libraries

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

## Analysis Code of Data

```{r}
#Read data into R 
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1") 

#Finding mean catch per region:

mean_region <- catch_original %>% #Create new variable 'mean region' as catch original, starting new pipline 
  select(-All, -notesRegCode) %>% #Then filter data to get rid of all and notes section
  mutate(Chinook = if_else(Chinook =='I', '1', Chinook)) %>% #Mutate Chinook values so that the letter I appears as the number 1
  mutate(Chinook = as.numeric(Chinook)) %>% #Makes the Chinook column numeric 
  pivot_longer(-c(Region, Year), #Pivot the datatable longer and keeping year and region the same 
               names_to = "species", #renames all columns with species name to just species 
               values_to ="catch") %>% #catch numbers in new column called catch
  mutate(catch = catch*1000) %>% #Change the value of catch to a mathmatical arrangement 
  group_by(Region) %>% #Grabbing region from datatable 
  summarize(mean_catch = mean(catch)) %>% #New variable mean catch is the mean of the catch data 
  arrange(desc(mean_catch)) #Arrange from most to least 
head(mean_region) #Display first few calculations of new table 
```

## Summary of Data

After the filtering of the data, we are able to see how many fish were caught in each region of the Alaska commercial fishing area between the years 1886-1997. This was done by coding through R Studio and cleaned up with tidy data techniques. The most fish were caught in region SSE, totaling over 3 million different types of salmon caught in that region.
