---
title: "a05"
format: html
author: "Sam Samler"
date: "04/01/2026"
---

# Introduction

This data set is sourced from University of California, Davis to study the socio-ecological trends around restoration lands. This data set explores the amount of visits restoration sites receive in and around the Bay Area as well as the visitor type recorded at the restoration site. The purpose of this study was to evaluate how humans respond to restored ecological sites in terms of their activity at a particular restored site. This data was collected by monitoring boat and car activity around restoration sites to get daily visit counts during daylight hours.

Below, I will clean and analyze the data to provide well organized charts and site mapping for the restoration areas based on the activity around the particular site based on my personal preferences for plotting data. \# Entering Code

## Installing Packages/Loading Libraries

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(forcats) # makes working with factors easier
library(ggplot2)
library(leaflet) # interactive maps
library(DT) # interactive tables
library(scales) # scale functions for visualization
library(janitor) # expedite cleaning and exploring data
library(viridis) # colorblind friendly color palette
```

## Entering Data

```{r}
delta_visits_raw <- read_csv('../data/processed/socioecological_monitoring_data.csv')
```

## Exploring Data

```{r}
colnames(delta_visits_raw)
```

```{r}
glimpse(delta_visits_raw)
```

```{r}
range(delta_visits_raw$Date)
```

```{r}
unique(delta_visits_raw$Time_of_Day)
```

## Cleaning Data

### Cleaning Column Names

```{r}
delta_visits <- delta_visits_raw %>% 
  janitor::clean_names() ## Making all names lowercase 
```

```{r}
colnames(delta_visits)
```

### Pivoting Longer for ggplot2

```{r}
visits_long <- delta_visits %>% 
  pivot_longer(cols = c(sm_boat, med_boat, lrg_boat, bank_angler, scientist, cars),
               names_to = 'visitor_type',
               values_to = 'quanitiy') %>% 
  rename(restore_loc = eco_restore_approximate_location) %>% 
  select(-notes)

head(visits_long)
```

### Calculating Daily Visits

```{r}
daily_visits_loc <- visits_long %>% 
  group_by(restore_loc, date, visitor_type) %>% 
  summarise(daily_visits = sum(quanitiy))
head(daily_visits_loc)
  
```

## Plot and Map Selection

### Plot Selection

```{r}
## Summarize total visits by restoration location 
daily_visits_totals <- daily_visits_loc %>% ## Then 
    group_by(restore_loc) %>% ## group by restoration location, then 
    mutate(n = sum(daily_visits)) %>% #create new column that contains total visits per location 
    ungroup() %>% ## remove grouping structure 
    mutate(restore_loc = fct_reorder(restore_loc, desc(n))) ## arrange total visits in descending order per restoration location
 
## Create faceted bar plot           
facet_plot <- ggplot(data = daily_visits_totals, ## Setting dataset 
       aes(x = visitor_type, y = daily_visits, ## Setting x and y 
           fill = visitor_type)) + 
    geom_col() + # Create bar chart 
    facet_wrap(~restore_loc, #Create separate pannels for resortation loaction 
               scales = "free_y", ## Each facet gets its own y-axis 
               ncol = 5, # number of columns 
               nrow = 2) + # Number of rows 
    scale_fill_viridis_d() + # Viridis for color visibility 
    labs(x = "Type of visitor", #Labeling x 
         y = "Number of Visits", #labeling 7
         title = "Total Number of Visits to Delta Restoration Areas", #setting titile 
         subtitle = "Sum of all visits during study period") + #subtitle 
    theme_bw() + 
    theme(legend.position = "bottom", #legend at bottom of plot 
          axis.ticks.x = element_blank(), #removing ticks 
          axis.text.x = element_blank()) #removing labels 

#Displaying plot 
facet_plot
```

### Map Selection

```{r}
## Creating new variable 
locations <- visits_long %>%
    distinct(restore_loc, .keep_all = TRUE) %>% # keeping only distinct restoration locations 
    select(restore_loc, latitude, longitude) #columns relevany for mapping 

#Creating Interactive Map 
leaflet(locations) %>%
    addProviderTiles(providers$Esri.WorldImagery) %>% #Base map, satellite 
    addProviderTiles(providers$CartoDB.PositronOnlyLabels) %>% #Overlay labels
    addCircleMarkers( #adding markers per restoration locations 
        lng = ~ longitude, # per longitude and 
        lat = ~ latitude, #per latitude 
        popup = ~ restore_loc, #showing names when location is clicked 
        radius = 5,#size of marker 
        # set fill properties
        fillColor = "salmon", #color of marker 
        fillOpacity = 1, #transparency of marker 
        # set stroke properties
        stroke = TRUE,
        weight = 0.5,
        color = "white",
        opacity = 1)
```

# Discussion

Based on the charts graphed above as well as the map generated using the longitudinal and latitudinal coordinates from the data set, it appears that most individuals are accessing the restoration sites via boat. The size and type of boat varies, however, it appears that is the main method of accessing the site. The North Delta is one of the only sites accessed mostly by car as well as Sherman Island, which makes sense as they are close to land compared to other sites on the map. Medium sized boats seems to be the main method of transportation to the restoration sites.

The purpose of the chart I chose was to display the type of visitor per site and how many visits each visitor type logged. I chose to do separate bar charts because I believe it is easier to read the graphs this was as opposed to them being smashed on one chart. I chose the satellite map, mostly because I thought it was the coolest option.
